---
title: "R Notebook"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
library(tidyverse)
library(patchwork)

set.seed(2022)
```

Read in toxin pres/abs files
```{r}
paths <- system("ls data/toxin_pres_abs/*.csv", intern = TRUE)

path <- paths[1]

read_toxin_pres_abs <- function(path){
 toxin <- path %>% basename() %>% str_remove("\\.csv")
  pa_df <- read_csv(path, show_col_types = FALSE) %>% 
    mutate(toxin = toxin) %>% 
    pivot_longer(-c(species, toxin),names_to = "toxin_gene", values_to = "pres_abs") %>% 
    dplyr::rename(genome = "species")
}

toxin_pres_abs <- map_df(paths, read_toxin_pres_abs) %>% 
  filter(!toxin %in% c("jam", "oci", "pag"))
```

Can find the historic markers here:
/for_laura/otu_analysis_curated

Don't stratify by individual toxin genes, either pure if all genes absent or a gene present in all genomes.

Read in clustering files

```{r}
otu_files <- system("ls data/otus/*/*.otu", intern = TRUE) %>% 
  data.frame(path = .) %>% 
 bind_cols(., unglue::unglue_data(.$path, "data/otus/{type}/{gene}.phylip.an.0.{dissim}.otu")) %>% 
  mutate(dissim = as.numeric(dissim),
         sim = 100 - dissim)

read_otus <- function(path){

col_names <- c("otu", "genome")

test <- read_tsv(path,col_names = col_names, show_col_types = FALSE) %>% 
  separate_longer_delim(genome, delim = ",") %>% 
  mutate(genome = str_remove(genome, ";.*"),
         path = path) %>% 
  left_join(otu_files) %>% 
  mutate(gene = str_remove(gene, "_curated"))
}

otus <- map_df(otu_files$path, read_otus)
```


Combine OTUs & toxin pres/abs data
```{r}

n_genomes_for_markers <- otus %>% 
  select(gene, genome) %>% distinct() %>% 
  group_by(gene) %>% 
  summarise(n_genomes_containing = n())

n_toxin_genes <- toxin_pres_abs %>% 
  select(toxin, toxin_gene) %>% 
  distinct() %>% 
  group_by(toxin) %>% 
  summarise(n_toxin_genes = n())

otu_toxins <- toxin_pres_abs %>% 
  left_join(otus)

otu_purity <- otu_toxins %>% 
  group_by(otu, toxin, gene, sim) %>% 
  mutate(mean_pres = mean(pres_abs),
         genomes_in_otu = n()) %>% 
  left_join(n_genomes_for_markers) %>% 
  left_join(n_toxin_genes)

genomes_in_pure_otus <- otu_purity %>% 
  group_by(toxin, gene, sim) %>% 
  filter(mean_pres == 0 | mean_pres == 1) %>% 
  summarise(percent_genome_in_pure_otu = ((n() / n_genomes_containing) / n_toxin_genes) * 100) %>% 
  distinct()

genomes_in_pure_otus %>% 
  group_by(gene, sim) %>% 
  summarise(avg_percent_in_pure_otus = mean(percent_genome_in_pure_otu)) %>% 
  ggplot(aes(sim, avg_percent_in_pure_otus, color = gene)) + 
  geom_point()

```

previously was considering each toxin gene independently, to match how this was done previously, first summarize pres/abs at toxin level (any gene hit = toxin present)

```{r}
toxin_otu_purity <- otu_toxins %>% 
  group_by(otu, toxin, sim, genome, gene, type) %>% 
  summarise(pres_abs = as.numeric(sum(pres_abs) > 0)) %>% 
  group_by(otu, toxin, sim, gene) %>% 
  mutate(mean_pres = mean(pres_abs),
         genomes_in_otu = n()) %>% 
  left_join(n_genomes_for_markers) 

toxin_genomes_in_pure_otus <- toxin_otu_purity %>% 
  group_by(toxin, gene, sim, type) %>% 
  filter(mean_pres == 0 | mean_pres == 1) %>% 
  summarise(percent_genome_in_pure_otu = ((n() / n_genomes_containing)) * 100) %>% 
  distinct()

toxin_genomes_in_pure_otus %>% 
  group_by(gene, sim) %>% 
  summarise(avg_percent_in_pure_otus = mean(percent_genome_in_pure_otu)) %>% 
  ggplot(aes(sim, avg_percent_in_pure_otus, color = gene, group = gene)) + 
  geom_point() + 
  geom_path()

plt_data_percent_in_pure <- toxin_genomes_in_pure_otus %>% 
  group_by(gene, sim, type) %>% 
  summarise(avg_percent_in_pure_otus = mean(percent_genome_in_pure_otu)) %>% 
  mutate(type = case_when(type == "best_amplicons" ~ "Amplicon from core genes",
                          type == "best_core_genes" ~ "Single copy core gene",
                          type == "historic" ~ "Traditional marker",
                          .default = type))
  
plot_order <- plt_data_percent_in_pure %>% 
  filter(sim == 99) %>% 
  dplyr::select(gene, plt_order = "avg_percent_in_pure_otus") %>% 
  arrange(plt_order)

colors3 <- c("lightblue3", "#194B77", "#826489")

plt_data_percent_in_pure %>% 
  left_join(plot_order) %>% 
  ggplot(aes(avg_percent_in_pure_otus, factor(gene,levels = plot_order$gene, ordered = TRUE), color = type, shape = factor(sim), label = round(avg_percent_in_pure_otus,1))) + 
  geom_point() + 
  geom_text(data = . %>% filter(sim == "99"),size = 2.5,nudge_x = 7) +
  scale_color_manual(values = colors3) +
  theme_bw() +
  labs(y = NULL, 
       x = "Average % of Total Toxin Sequences in Pure OTUs", 
       shape = "OTU % identity",
       color = "Marker type")

ggsave("results/total_percent_toxin_seqs_in_pure_OTUs.png", width = 4, height =3, scale = 1.5)
ggsave("results/total_percent_toxin_seqs_in_pure_OTUs.pdf", width = 4, height =3, scale = 1.5)
ggsave("results/total_percent_toxin_seqs_in_pure_OTUs.eps", width = 4, height =3, scale = 1.5)

```


```{r}

change_toxin_names <- function(short_name) {
 case_when(short_name == "aer" ~ "Aeruginosin",
           short_name == "apn" ~ "Anabaenopeptin",
           short_name == "jam" ~ "Jamaicamides",
           short_name == "mcn" ~ "Cyanopeptolin",
           short_name == "mcy" ~ "Microcystin",
           short_name == "mdn" ~ "Microviridin",
           short_name == "oci" ~ "Cyanopeptin",
           short_name == "pag" ~ "Prenylagaramide",
           short_name == "pir" ~ "Piricylamide"
           ) 
}


plt_data_percent_in_pure <- toxin_genomes_in_pure_otus %>% 
  group_by(gene, sim, toxin, type) %>% 
  summarise(avg_percent_in_pure_otus = mean(percent_genome_in_pure_otu)) %>% 
  mutate(type = case_when(type == "best_amplicons" ~ "Amplicon from core genes",
                          type == "best_core_genes" ~ "Single copy core gene",
                          type == "historic" ~ "Traditional marker",
                          .default = type),
         toxin = change_toxin_names(toxin))
  
plot_order <- plt_data_percent_in_pure %>% 
  ungroup() %>% 
  filter(sim == 99) %>% 
  dplyr::select(gene, plt_order = "avg_percent_in_pure_otus") %>% 
  group_by(gene) %>% 
  slice_max(plt_order, n = 1) %>% 
  arrange(plt_order) %>% ungroup() %>% distinct()

plot_order <- plt_data_percent_in_pure %>% 
  ungroup() %>% 
  filter(sim == 99) %>% 
  dplyr::select(gene, plt_order = "avg_percent_in_pure_otus") %>% 
  group_by(gene) %>% 
  summarise(plt_order = mean(plt_order)) %>% 
  arrange(plt_order) %>% ungroup() %>% distinct()


facet_order <- plt_data_percent_in_pure %>% 
  ungroup() %>% 
  filter(sim == 99) %>% 
  dplyr::select(gene, avg_percent_in_pure_otus, toxin) %>% 
  group_by(toxin) %>% 
  summarise(mean_pure= mean(avg_percent_in_pure_otus)) %>% 
  arrange(desc(mean_pure)) %>% 
  pull("toxin")
  

colors3 <- c("lightblue3", "#194B77", "#826489")

plt_data_percent_in_pure %>% 
  left_join(plot_order) %>% 
  ggplot(aes(avg_percent_in_pure_otus, factor(gene,levels = plot_order$gene, ordered = TRUE), color = type, shape = factor(sim), label = round(avg_percent_in_pure_otus,1))) + 
  geom_point() + 
  theme_bw() +
  facet_wrap(~factor(toxin,levels = facet_order, ordered = TRUE), nrow = 2) +
  scale_color_manual(values = colors3) +
  theme_bw() +
  labs(y = NULL, 
       x = "Average % of Total Toxin Sequences in Pure OTUs", 
       shape = "OTU % identity",
       color = "Marker type")

ggsave("results/per_toxin_percent_in_pure_OTUs.png", width = 3.5, height =3.5, scale = 2.5)
ggsave("results/per_toxin_percent_in_pure_OTUs.pdf", width = 3.5, height =3.5, scale = 2.5)
ggsave("results/per_toxin_percent_in_pure_OTUs.eps", width = 3.5, height =3.5, scale = 2.5)
```






```{r}
file.name1 <-"mdn matrix"
mdn<- read.csv(paste0(file.name1,".csv"), header=T)
options(max.print = 10000)
apn%>%
  filter(species %in% c("Microcystis_aeruginosa_PCC_7806SL","Microcystis_aeruginosa_Ma_AC_P_19900807_S300",
  "Microcystis_aeruginosa_DIANCHI905"))

```

```{r}
###OTU shared presence per toxin graph (dont use)###
library(ggplot2)
file.name1 <-"OTU graph data"
dat<- read.csv(paste0(file.name1,".csv"), header=T)
ggplot(data= dat, aes(OTU.id, shared.presence, group = Gene, color = Gene, shape = Gene))+
  #geom_line()+
  geom_point()+
  scale_shape_manual(values = seq(0,8))+
  labs(shape = "Marker Gene", color= "Marker Gene", x = "OTU % ID", y = "Shared Toxin Gene Presence (%)" )+
    facet_wrap( ~ toxin, nrow = 5)

###OTU purity per toxin###
library(ggplot2)
file.name1 <-"OTU graph data purity"
dat<- read.csv(paste0(file.name1,".csv"), header=T)
ggplot(data= dat, aes(x=factor(OTU.id), shared.presence, group = Gene, color = Gene, shape = Gene))+
  #geom_line()+
  geom_point()+
  scale_x_discrete(labels=c("95"="95","97"="97","99"="99"))+
  scale_shape_manual(values = seq(0,8))+
  labs(shape = "Marker Gene", color= "Marker Gene", x = "OTU % ID", y = "% Sequences in Pure OTUs")+
  facet_wrap( ~ toxin, scales = "free_x", nrow = 5)

###OTU purity per gene for all toxins summary###
file.name2 <-"OTU summary"
dat2<- read.csv(paste0(file.name2,".csv"), header=T)
ggplot(data= dat2, aes(x=factor(OTU.id), shared.presence, color = Gene, shape = Gene))+
  geom_point()+
  scale_x_discrete(labels=c("95"="95","97"="97","99"="99"))+
  scale_shape_manual(values = seq(0,8))+
  labs(shape = "Marker Gene", color= "Marker Gene", x = "OTU % ID", y = "Average % of Total Toxin Sequences in Pure OTUs")

###toxin gene ref table###
library(kableExtra)
file.name1 <-"Toxin ref sequences"
dat<- read.csv(paste0(file.name1,".csv"), header=T)
dat %>%
  kbl(caption = "Table 2: Toxin sequences used in Blast query of Microcystis genomes from Supplemental Table 1 to generate presence absence matrix of toxin genes.") %>%
  kable_classic_2(full_width = F, html_font = "Times New Roman")%>%
  kable_styling(font_size = 12)

###microcystis genome ref table###
file.name3 <-"microcystis genome ref"
dat3<- read.csv(paste0(file.name3,".csv"), header=T)
kbl(dat3,longtable = T, booktabs = T, caption = "Table 1: Published Microcystis Genomes") %>%
  kable_classic_2(full_width = F, html_font = "Times New Roman")%>%
  kable_styling(font_size = 12, latex_options = c("repeat_header"))%>%
  save_kable(file = "D:/Ph.D/Dissertation/Prelim Project- Microcystis Strain Variation Markers/Figure 4 OTU/mcy_table.pdf")

```

