---
title: "Read duplication"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
```


Read QC and MG-RAST report a high level of sequence duplication. However, the details of this are obscured. Here, I'm looking at this duplicaiton in more detail. Specifically I'd like to know how duplicated the most duplicated reads are, how many reads are duplicated, and what some of the duplicated reads align to (manually BLASTing the top few duplicated sequences).

```{r}
library(Biostrings)
library(tidyverse)
```
Read in the forward reads
```{r}
f_reads_qc <- Biostrings::readDNAStringSet("data/qc_sequence_files/S3_Fallen_R1.fastq.gz",format = "fastq") %>% as.data.frame() %>% 
  rownames_to_column("seq_id") %>% 
  rename(seq = "x")
```


How many reads are unique?
```{r}
unique_reads <- f_reads_qc %>% distinct(seq)
percent_unique <- nrow(unique_reads) / nrow(f_reads_qc) * 100
```

Are duplicated reads only duplicted a few times, or many?
```{r}
duplicate_count <- f_reads_qc %>%
  group_by(seq) %>% 
  summarise(n_duplicates = n())

duplicate_count %>% 
  ggplot(aes(n_duplicates)) +
  scale_y_log10(breaks = c(1,10,100,1000,10000,10000,100000), labels = c("1","10","100","1,000","10,000","100,000","1,000,000")) +
  geom_histogram() +
  geom_vline(xintercept = max(duplicate_count$n_duplicates),color = "red")
```

All duplicate raw sequences of the most duplicated sequence
```{r}
seq_most_repeated <- duplicate_count %>% top_n(1,n_duplicates) %>% pull(seq)

most_repeated_obs <- f_reads_qc %>% filter(seq == seq_most_repeated)
```



